---
title: "Add Standard Groupings"
description: >
  Add variables with standard PSRC category groupings
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Add Standard Groupings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Agency standard category schemes

In light of sample sizes, breakdowns from PSRC's Household Travel Survey instrument often involve category groupings with less than the most granular reported level of detail. **psrc.travelsurvey** functions provide the most commonly utilized groupings, which can be considered agency standard for the variables of interest. Although each new variable is specific to one unit of analysis (i.e. hh, person, day, trip, vehicle), the first and only argument to each grouping function is the entire `hts_data` object, in order to facilitate piping. When more than one standard grouping exists for a variable, the function returns them all, for convenience; they can be distinguished by a suffix denoting the number of categories.

```{r example, message=FALSE, eval=TRUE}
library(psrc.travelsurvey)
library(magrittr)
library(data.table)

# Specify which variables to retrieve
vars <- c("hhincome_broad", "age", "employment", "education", "hhsize", 
          "mode_characterization", "distance_miles", "duration_minutes")

# Retrieve the data
hts_data <- get_psrc_hts(survey_vars = vars) 

# Add standard groupings for age and household size
# -- Notice age and hhsize use different tables (person, hh)
# -- but you don't need to worry about that; the package knows
hts_data <- hts_data %>% 
  hts_bin_age() %>%
  hts_bin_hhsize()

hts_data$person[, head(.SD), .SDcols=patterns("^age")]

```

These functions each require a disaggregate input variable to be requested in the earlier `get_psrc_hts()` call. Both the input variable and the output variable are included in the resulting `hts_data` object.

Currently, the following functions exist, along with the required input variable.


|           Function               |  Input variable   |    Output variable(s)   |                 Breakpoints                |
|:---------------------------------|:-----------------:|:-----------------------:|:------------------------------------------:|
| **`hts_bin_income()`**           |  hhincome_broad   |  hhincome_bin5          |  $25K ]( $50K ]( $75K ]( $100K             |
| **`hts_bin_hhsize()`**           |  hhsize           |  hhsize_bin4            |  1 • 2 • 3 • 4+                            |
| **`hts_bin_vehicle_count()`**    |  vehicle_count    |  vehicle_count_bin4     |  1 • 2 • 3 • 4+                            |
| **`hts_bin_age()`**              |  age              |  age_bin3/bin5          | 18 )[ 64 )[ 65 **or** 18 )[ 25 )[ 45 )[ 65 |
| **`hts_bin_worker()`**           |  employment       |  worker                 |  worker • non-worker                       |
| **`hts_bin_edu()`**              |  education        |  education_bin2         |  Less )[ Bachelors+                        |
| **`hts_bin_gender()`**           |  gender           |  gender_bin3            |  m • f • nb/oth                            |
| **`hts_bin_sexuality()`**        |  sexuality        |  sexuality_bin3         |  hetero • LGBTQ+ • Missing                 |
| **`hts_bin_commute_freq()`**     |  commute_freq     |  commute_freq_bin6      |  Less • 1 • 2 • 3 • 4 • 5+ days/week       |
| **`hts_bin_commute_mode()`**     |  commute_mode     |  commute_mode_bin5      |  SOV • HOV • trnst • w/b • oth             |
| **`hts_bin_lum_sector()`**       |  industry         |  lum_sector             |  (13 sectors)                              |
| **`hts_bin_industry_sector()`**  |  industry         |  industry_sector        |  (8 sectors)                               |
| **`hts_bin_telecommute_freq()`** |  telecommute_freq |  telecommute_freq_bin4  |  Less )[ 1 )[ 3 )[ 5+                      |
| **`hts_bin_telework_time()`**    |  telework_time    |  telework_time_bin3     |  Less )[ 3 )[ 6 hrs                        |
| **`hts_bin_dest_purpose()`**     |  dest_purpose     |  dest_purpose_bin4/bin9 |  Home • Work • Social/Rec/Eat • Errand     |
|                                  |                   |                         |                                            |






