---
title: "Retrieve and summarize sss data"
description: >
  Retrieve and summarize transportation safety and security survey data
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve and summarize sss data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
suppressWarnings(library(psrc.travelsurvey))
```

## For use with the 2025 PSRC Safety & Security Survey

The Safety & Security Survey is a more traditional attitudinal survey, with responses at a person level, rather than multiple related tables. (The **travelSurveyTools** package is not used in handling it.) Use `get_psrc_sss()` to retrieve the variables you need, then use `psrc_sss_stat()` to calculate weighted counts, shares, or numeric summaries.

Because the package queries Elmer, you'll need to be connected in office or through VPN to run the retrieval step.

## Data retrieval

Since psrc.travelsurvey uses Elmer, the agency's central database, you'll need to be connected in office or through VPN. Then use `get_psrc_sss()`, which takes a single argument:

  * **survey_vars** - a vector of desired survey variable names

The return object is a person-level `data.table`. `person_id`, `hh_id`, `sample_segment`, and `person_weight` are included automatically, so they do not need to be added to `survey_vars`.

```{r retrieve_sss, message=FALSE, eval=FALSE}
library(psrc.travelsurvey)
library(magrittr)
library(dplyr)

vars <- c("employment", "crash_participant", "travel_anxiety", "safety_choices", "crash_number_people")

sss_data <- get_psrc_sss(survey_vars = vars)
```

## Summarization

`psrc_sss_stat()` assumes a person-level analysis unit. 
The main arguments are:

  * **sss_data** - the table returned by `get_psrc_sss()`
  * **group_vars** - one or more grouping variables, in nesting order
  * **stat_var** (optional) - a numeric variable for min, max, median, and mean summaries

Like `psrc_hts_stat()`, it uses the last grouping variable as the share variable when `stat_var` is omitted.

### Count and share example

This example estimates the weighted distribution of `travel_anxiety` responses within each `crash_participant` category.

```{r sss_cat, message=FALSE, eval=FALSE}
rs1 <- psrc_sss_stat(
  sss_data,
  group_vars = c("crash_participant", "travel_anxiety"),
  incl_na = FALSE
)

head(rs1)
```

The resulting table can be read as an association summary: within each `crash_participant` group, compare the weighted shares across `travel_anxiety` response levels.

### Numeric summary example

TSS variables are primarily dichotomous or ordinal, but `psrc_sss_stat()` can also calculate handle weighted means for any numeric variables, using the `stat_var` argument.

```{r sss_num, message=FALSE, eval=FALSE}
sss_data <- mutate(sss_data, crash_number_people_numeric = if_else(
    is.na(crash_number_people), NA_integer_, as.integer(stringr::str_extract(crash_number_people, "^\\d+"))))

rs2 <- psrc_sss_stat(
  sss_data,
  group_vars = "crash_participant",
  stat_var = "crash_number_people_numeric",
  incl_na = FALSE
)

head(rs2)
```

## Notes

Margin-of-error columns are returned where the underlying survey calculation produces standard errors; if a statistic cannot support a variance estimate, its MOE field will be `NA`.