---
title: "Retrieve and summarize data"
description: >
  Retrieve and summarize household travel survey data
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieve and summarize data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Works with **travelSurveyTools**

The **psrc.travelsurvey** package is meant to complement and enhance the **travelSurveyTools** package for PSRC users. It delivers data in the `hts_data` format appropriate as input to subsequent **travelSurveyTools** functions. It also provides wrapper functions that provide PSRC defaults to **travelSurveyTools** functions, resulting in simpler workflows with fewer required arguments. **psrc.travelsurvey** is constrained to an extent by fitting into the **travelSurveyTools** system, but if you have suggestions for other features or functions you'd find helpful, please contact the authors and we'll see if we can implement them.

## Data retrieval

Since psrc.travelsurvey uses Elmer, the agency's central database, to retrieve PSRC household travel survey data you'll need to be connected (in office or via VPN). Then use `get_psrc_hts()`, which takes two arguments:

  * **survey_years**  - a vector of one or more years; the default is all available i.e., `c(2017,2019,2021,2023)` as of this writing.
  * **survey_vars** - a vector of variable names (regardless of the table in which they occur) 

The return object is a list with separate data.table elements for each potential unit of analysis, i.e. household (`hh`), `person`, `day`, `trip`, and `vehicle`. The **travelSurveyTools** package workflow uses this central object for related or independent variable summaries. It can include the full set of variables for all your summaries, so it's unnecessary to request separate data objects to feed each separate summary data table. It also supports multiyear summaries (trends/comparisons) via the `survey_year` field.
  
For efficiency, `get_psrc_hts()` only requests the variables you specify, so you must know their precise names. A lookup function, `psrc_hts_varsearch()` can assist; it takes a single string argument as the codebook search term and matches either the variable name or description fields. Regular expressions are supported for more complex searches; for example, the word boundary operator `\\b` can be used to omit vehicle model from results when searching for "mode".

```{r varsearch, message=FALSE, eval=FALSE}
View(psrc_hts_varsearch("\\bmode\\b"))
```

To browse the entire variable list, open the internal data object `init_variable_list`:
```{r init_variable_list, message=FALSE, eval=FALSE}
View(psrc.travelsurvey:::init_variable_list)
```

Note: It is unnecessary to include the survey year, table ids, weights, or survey stratification variable (`sample_segment`) in your `survey_vars` argument; since they are necessary for later steps, `get_psrc_hts()` includes them implicitly. Factor variables and levels are applied from the codebook, and missing or skip-value codes are recoded as `NA`.

## Summarization

To summarize data, use `psrc_hts_stat()`. This has four primary arguments, and one option. The arguments are:

  * **hts_data**  - the `hts_data` object
  * **analysis_unit** - string specifying the scale at which statistics are calculated, either "hh", "person", "day","trip", or "vehicle".
  * **group_vars** - any grouping variables, in nesting order (share summaries are calculated per categories of the variable listed last).
  * **stat_var** (optional) - Numeric variable if min/max/median/mean is desired; omit for count/share summaries.

As an example, we'll retrieve the following survey variables, then do one count/share summary, and one numeric summary. Notice, travelSurveyTools handles the relational element of the data, so tables lower in the organizational heirarchy (like `trips`) inherit attributes from those above them (such as `hh`); this is a handy feature and avoids unnecessary data duplication among tables.

```{r retrieve_and_summarize, message=FALSE, eval=TRUE}
library(psrc.travelsurvey)
library(data.table)

# Specify which variables to retrieve
vars <- c("hhsize", "employment", "commute_mode", "dest_purpose",
          "mode_characterization", "distance_miles")

# Retrieve the data
hts_data <- get_psrc_hts(survey_vars = vars)  # default includes all survey_years
  
# Calculate a numeric summary, i.e. min/max/median/mean
rs1 <- psrc_hts_stat(hts_data, 
                     analysis_unit="trip", 
                     group_vars="hhsize", 
                     stat_var="distance_miles")

head(dplyr::select(rs1[survey_year==2023], -c(survey_year, min, max)))

# Calculate a categorical summary, i.e. count/share
rs2 <- psrc_hts_stat(hts_data, "trip", c("hhsize", "mode_characterization"))

head(dplyr::select(rs2[survey_year==2023 & as.character(mode_characterization)=="Transit"],
                   -c(survey_year, mode_characterization)))
```

## Custom variables and filtering

Since each `hts_data` element is a data.table, they can be manipulated with any combination of base r, tidyverse (e.g. `mutate()`), or data.table (e.g `:=`) syntax. The example below illustrates adding variables specific to the `hts_data$trip` element. Notice, the inheritance capability mentioned earlier applies to custom variables as well, so if you want to recode a household-level variable to feature in a trip-level analysis, just add it to `hts_data$hh` (not to `hts_data$trip`).  

Standard category groupings not present in the base data can be added via the functions described in [the accompanying vignette](add-standard-groupings.html). These take the entire `hts_data` object as argument, not a specific table (the package knows which is involved). This lends itself to piping multiple standard recodes.

Keep in mind that filtering any table removes records available for subsequent analyses. Rather than manage multiple filtered copies of the data object, you may want to create a new variable on the relevant table with `NA` values for the records you intend to exclude from the summary. You can then use the `incl_na=FALSE` option to exclude that line from summary results (of particular value when you want shares calculated from only reported categories). Here we will add a variable specific to work-related trips, and use the `incl_na=FALSE` option to limit the statistics (including percentages) to work-related trips only, without erasing non-work trips from the `hts_data` object or creating a separate `hts_data` object for this summary. 

```{r add_variable, message=FALSE, eval=TRUE}
library(dplyr)
library(stringr)

# Use a standard variable recode to get simpler trip purpose
hts_data <- hts_bin_dest_purpose(hts_data)

# Use dplyr::mutate to add a simplified mode field & work purpose specifier
hts_data$trip <- mutate(
  hts_data$trip, 
  purpose_work = case_when(
    dest_purpose_bin4=="Work" ~ "Work",
    !is.na(dest_purpose_bin4) ~ NA_character_),
  mode_basic = case_when(
    mode_characterization=="Airplane"                 ~NA_character_,
    str_detect(mode_characterization, "HOV")          ~"HOV2+",
    mode_characterization=="Drive SOV"                ~"Drive alone",
    str_detect(mode_characterization, "^(Walk|Bike)") ~"Walk/Bike/Micromobility",
    TRUE ~mode_characterization)
) 

# Calculate a categorical summary, i.e. count/share
rs3 <- psrc_hts_stat(hts_data, "trip", c("purpose_work", "mode_basic"), incl_na=FALSE)

head(rs3[survey_year==2019])

```

Note: **travelSurveyTools** summary functions require additional preparatory steps that **psrc.travelsurvey** handles for you, for simplicity. To use native **travelSurveyTools** summary functions, you'll need to handle these extra steps yourself, since **travelSurveyTools** does not have a dependency on the **psrc.travelsurvey** package.
