---
title: "Summarize trip and VMT rates"
description: >
  Specific functions for average daily trip count and vehicle miles traveled
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarize trip and VMT rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
suppressWarnings(library(psrc.travelsurvey))
```

## Why rates? A convenient way to examine and compare

In addition to [aggregate counts and shares](retrieve-and-summarize-data.html#summarization), **psrc.travelsurvey** includes functions to calculate average daily trip- and VMT (vehicle miles traveled) rates. These metrics are often used in travel analysis and modeling. As a ratio (mean amount per person per day), these rates are normalized in respect to both population and days reported, while maintaining a unit scale so they can be interpreted or compared in both absolute and relative terms. (In contrast, shares have no unit and can only be compared in relative terms; counts aren't normalized and by themselves can't be meaningfully compared across subgroups or surveys.)

## psrc.travelsurvey functions

Trip and VMT rates involve summarization at the day level prior to aggregate summarization across respondents, and have their own functions:

  * **`psrc_hts_triprate()`**
  * **`psrc_hts_vmtrate()`**

Each function is simplified from `psrc_hts_stat()`; since `analysis_unit` and `stat_var` for each are fixed/inherent, there are only two arguments:

  * **hts_data**  - the `hts_data` object
  * **group_vars** - any grouping variables, in nesting order (share summaries are calculated per categories of the variable listed last).
  
The `incl_na=FALSE` option is also available in both functions.
  
### Requirements

These functions rely on additional functionality submitted to the **travelSurveyTools** package that has not yet been merged, so you'll need to install a branch of that package: 

    `remotes::install_github("RSGInc/travelSurveyTools", ref="add_vmt_rate")`

Because VMT is calculated from distance, is limited to vehicular travel, and accounts for co-travelers, the preceding `get_psrc_hts()` call must include the following among the `survey_vars` argument:

  * **`mode_class`**
  * **`trip_distance`**
  * **`travelers_total`**. 
  
If the hts_data object is missing one or more of these, **`psrc_hts_vmtrate()`** will report that to the user and return NULL.

Because count is an implicit trip attribute, **`psrc_hts_triprate()`** requires no additional variables.

### Examples

```{r triprate, message=FALSE, eval=TRUE}
library(psrc.travelsurvey)
library(data.table)

# Specify which variables to retrieve
vars <- c("telecommute_freq", "workplace",
          "mode_class", "distance_miles", "travelers_total")

# Retrieve the data
hts_data <- get_psrc_hts(survey_vars = vars)  # default includes all survey_years
hts_data <- hts_bin_telecommute_trichotomy(hts_data)

# Calculate a trip rate
rs1 <- psrc_hts_triprate(hts_data, group_vars="telecommute_trichotomy")

head(dplyr::select(rs1[survey_year==2023], -c(survey_year, count, min, max, median)))

# Calculate a VMT rate
rs2 <- psrc_hts_vmtrate(hts_data, "telecommute_trichotomy", incl_na=FALSE)

head(dplyr::select(rs2[survey_year==2023], -c(survey_year, count, min, max, median)))
```
